---
title: "Reproducible Research: Peer Assessment 1"
author: "AlekseyVosk"
date: '20 may 2018 '
output: 
  html_document:
    keep_md: true
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
Sys.setlocale("LC_TIME", "C")
```

For research we will need functions from external packages.
```{r}
library(tidyverse)
```

## Loading and preprocessing data
Read csv file into variable "data".
```{r}
data <- read_csv("activity.csv", col_types = "iDi")
```
Check, if there any NAs in data.
```{r}
apply(apply(data, 2, is.na), 2, sum)
```
There are 2304 missing values in the column "steps" in the dataset.
For the further analysis let's make dataframe with additional column "f_interval", which represent intervals as factor variables, and dataframe with no missing values.
```{r}
mod_data <- data %>%
  mutate(f_interval = as.factor(interval))
complete_data <- filter(mod_data, !is.na(steps))
```

## What is mean total number of steps taken per day?
We will work with dataframe with no missing values, and make new dataframe, that represent total number of steps taken per day.
```{r}
steps_per_day <- complete_data %>%
  group_by(date) %>%
  summarise(sum = sum(steps))
print(steps_per_day, n=3)
```
To visualise this dataframe let's make histogramm of total number of staps taken per day:
```{r}
steps_per_day_ggplot <- ggplot(data = steps_per_day, mapping = aes(x = sum)) + 
  geom_histogram(fill = "blue", alpha = 0.3, binwidth = 1000) +
  labs(x = "Total number of steps", y = "Number of days, count", title = "Total number of steps taken each day")
print(steps_per_day_ggplot)
```

Mean and median of total number of staps taken per day:
```{r}
mean_steps_per_day <- mean(steps_per_day$sum)
median_steps_per_day <- median(steps_per_day$sum)
print(paste("Mean:", round(mean_steps_per_day, 1), " ", "Median:", median_steps_per_day))
```

## What is the average daily activity pattern?
We will work with dataframe with no missing values, and make new dataframe, that represent mean number of steps taken per interval.
```{r}
steps_per_interval <- complete_data %>%
  group_by(f_interval) %>%
  summarise(mean = mean(steps))
print(steps_per_interval, n=3)
```
To visualise this dataframe let's make time series plot of mean number of staps taken per interval:
```{r}
steps_per_interval_ggplot <- ggplot(data = steps_per_interval, mapping = aes(x = f_interval, y = mean, group = 1)) + 
    geom_path(colour = "red") +
    labs(x = "Interval of time", y = "Number of steps, mean", title = "Mean number of steps taken in each 5-minute interval")
print(steps_per_interval_ggplot)
```

Interval, which contains, on average, the maximum number of steps:
```{r}
interval_with_max_steps <- as.character(arrange(steps_per_interval, desc(mean))[["f_interval"]][1])
print(paste("Interval of maximum:", interval_with_max_steps))
```
So, it is interval, which corresponds to 8:35 am.

## Imputing missing values
Total number of missing values in the dataset:
```{r}
sum(!complete.cases(data))
```
As we previously saw, all missing valius in dataset is in colums "steps".
Lets make new dataframe with colums "step_filled", in which all the values are same as in "steps", except NAs are replaced by the mean values of steps in the same time interval across all dataset.  
First let's make dataframe with mean values of steps:
```{r}
mean_steps_per_interval <- complete_data %>%
  group_by(f_interval) %>%
  summarise(mean = mean(steps))
print(mean_steps_per_interval, n=3)
```
Then let's apply replacement function to the "steps" column and make new dataframe "data_filled".
```{r}
step_fill <- function (step, inter) {
  if (is.na(step)){
    mean_steps_per_interval[mean_steps_per_interval$f_interval == inter,][[2]]
  }
  else step
}

data_filled <- mod_data %>%
  mutate(steps_filled = mapply(step_fill, steps, f_interval))
print(data_filled, n=3)
```
Now let's make histogramm of total number of staps taken per day, based on our new "steps_filled" value:
```{r}
steps_per_day_filled <- data_filled %>%
  group_by(date) %>%
  summarise(sum = sum(steps_filled))

steps_per_day_filled_ggplot <- ggplot(data = steps_per_day_filled, mapping = aes(x = sum)) +
  geom_histogram(fill = "blue", alpha = 0.3, binwidth = 1000) +
  labs(x = "Total number of steps", y = "Number of days, count", title = "Total number of steps taken each day")
print(steps_per_day_filled_ggplot)
```

On histogramm we can see much more days with about 11000 steps taken.
```{r}
mean_steps_per_day_filled <- mean(steps_per_day_filled$sum)
median_steps_per_day_filled <- median(steps_per_day_filled$sum)
print(paste("New mean:", round(mean_steps_per_day_filled, 1), " ", "New median:", round(median_steps_per_day_filled)))
```
Mean and median almost didn't change, which is a good sign.

## Are there differences in activity patterns between weekdays and weekends?
Let's create new dataframe with factor variable in the dataset with two levels - "weekday" and "weekend"
```{r}
week_day_end <- function (day) {
  if (day == "Saturday" | day == "Sunday") {
    "weekend"
  } else {
    "weekday"
  }
}

steps_per_weekday <- data_filled %>%
  mutate(wday = weekdays(date), weekde = as.factor(map_chr(wday ,week_day_end))) %>%
  select(steps_filled, f_interval, weekde)
print(steps_per_weekday, n=3)
table(steps_per_weekday$weekde)
```
To illustrate differences in activity patterns between weekdays and weekends lets make a panel plot based on new "weekday"/"weekend" variable:
```{r}
steps_per_weekday_interval <- steps_per_weekday %>%
  group_by(f_interval, weekde) %>%
  summarise(mean_steps = mean(steps_filled))

steps_per_interval_filled_week_ggplot <- 
    ggplot(data = steps_per_weekday_interval, mapping = aes(x = f_interval, y = mean_steps, colour = weekde, group = 1)) +
    geom_path() +
    facet_wrap(~ weekde, nrow = 2) +
    labs(x = "Time of the day, 5 min intervals", y = "Number of steps, count", title = "Number of steps taken in weekdays and weekends") +
    guides(colour=FALSE)
print(steps_per_interval_filled_week_ggplot)
```

We can see some differences in activity patterns between weekdays and weekends.